{\includegraphics[width=3cm, height=2.5cm]{Cran.jpg}} \\ \includegraphics[width=7cm, height=1cm]{TB.jpg} \\ {Technical Documentation for TersectBrowser+}

### 2025-05-05 TechDoc/TechDoc.tex
upload feature search text

% Have put them in alphabetical order
\author{Gabrielle Baumberg, Gregory Lupton, David Oluwasusi, Tanya Stead \\ Supervised by: Dr. Tomasz Kurowski}

@@

\section{Overview}
This document describes the technical specifications and extended functionalities of Tersect Browser+, a web-based application for the exploration of genome and variant data in cultivated and wild-type crops. Building upon the original Tersect Browser, developed by Tomasz Kurowski [Reference], this enhanced version integrates additional features to support the identification and analysis of introgressions within accession species. The system leverages a MongoDB database to manage genomic datasets efficiently and provides a suite of interactive tools for biological interpretation.

The original Tersect Browser offered a streamlined interface for investigating genomic intervals and phylogenetic relationships, powered by the high-performance Tersect CLI. It included dynamic visualizations such as phylogenetic trees and genetic distance heatmaps derived from resequenced genome data.

Tersect Browser+ extends this foundation with several key enhancements aimed at cultivar biology. These include a genome browser, gene model visualizer, feature search tools, and barcode generation utilities—together offering a more powerful and versatile platform for crop genetics research.

The requirements for the application extension were described in [reference Tom’s briefing on this from March]. For the biologist to be able to:

\begin{itemize}
    \item open a genome browser within Tersect Browser and investigate specific variants.
    \item visualise gene models in areas of suspected introgressions.
    \item identify areas where high-impact variants are visible. 
    \item be able to genereate DNA barcodes for the varioius cultivars and species in the dataset. 
    \item to be able to load additional datasets, including for other crops. 
\end{itemize}

The extensions were designed and added by the named authors under the supervision of of Dr Tomasz Kurowski between March and May 2025.

%This Technical Documentation describes the particular details of building upon the original TersectBrowser developed by Tomasz Kurowski into  TersectBrowser+. It outlines the integration of multiple extension features, providing the user with additional display and data analysis capabilities.

\section{Architecture}
The \textit{Tersect Browser+} application is composed of four main components: 
\begin{itemize}
    \item the frontend; 
    \item the backend; 
    \item common utilities; and, 
    \item extension modules.
\end{itemize}

[Insert figure in here]

The frontend is a browser-based interface developed in Angular and React, designed for interactive data visualization, including distance plots, accession selectors, and genome browser integration. The backend, built in Node.js and Express, provides API endpoints, data models, and utility scripts to serve genomic data and computation results from a MongoDB database. A set of common modules supports both frontend and backend logic with shared resources. Additionally, the architecture supports pluggable extensions, such as the genome-browser, which provide domain-specific capabilities like rendering gene models or enabling barcode-based comparisons. Supporting data files, environment configurations, and Python scripting for preprocessing or bioinformatics workflows are organized in separate folders for data loading and virtual environments, enabling efficient pipeline integration and deployment.

%The TersectBrowser+ software is an Angular project, with an original frontend and backend written in Angular 8, and a separate extension space written in Angular 19 using Node 22. This `bolt-on' approach was chosen for the extensions in order to seamlessly integrate the new extensions with the original workspace within the project time constrainsts. Additionally, this approach allows the new extnesions to use up-to-date packages, resulting in enhanced functionality and a modern design. To run TersectBrowser+ on different platforms, the new extensions are hosted on NPM and can be linked as dependencies within the project.

%Extensions added to TersectBrowser+ include the integration of a genome browser capable of displaying a variety of different tracks, notably variants and gene models, search functionality to identify and display regions containing high impact variants, and a tool to generate accession-specific barcodes. These extensions are hosted in a separate section of the project, but interact with both frontent and backend services to provide a seamless user experience.

\textit{Tersect Browser+} is primarily built using Angular 8, preserving the architecture of the original application. To enable modern feature development without disrupting the existing codebase, a dedicated extension layer has been introduced using Angular 15 and Node.js 22. This modular, "bolt-on" extension framework allows for the seamless integration of new capabilities while taking advantage of the latest frontend and backend technologies. Extensions are developed independently and published to NPM, making them easily linkable through project dependencies. This hybrid approach ensures forward compatibility and scalability, while maintaining stability in the core application. Future versions of \textit{Tersect Browser+} may explore fully migrating the main application to a more unified and modern framework.

\section{Deployment}
The GitHub project \url{https://github.com/Tersect-Browser/Tersect-browser} README provides detailed instructions for setup, dependencies, and deployment. This section provides additional information.

%Here, we expand upon these instructions and give a detailed overview of the structure of each extension.

@@

The Admin is defined as an individual with bioinformatics and software design knowledge, background comprehension for all sections of this Technical Documentation, and acting to facilitate the use of a deployed TersectBrowser+ by the plant breeder main user of the software. When a new dataset is requested for deployment on TersectBrowser+, the Admin will need to follow the below steps:
\begin{enumerate}
    \item Collect background context and required data from the user.
\begin{itemize}
    \item Reference genome in fasta format (may be compressed). Size and chromosome number will impact the speed of dataset upload and deployment - TersectBrowser+ has been tested with a genome size of ~1GB, and 20 chromosome pairs.
    \item Resequenced genome dataset in VCF format (may be compressed, as a multi-sample VCF or as a directory of individual files). TersectBrowser+ has been tested with a dataset size of ~500 VCF files representing individual accessions. VCF files should be produced using SNPeff software. If not using SNPeff, ensure that variant impact levels are categorised into `High', `Moderate', and `Low' impact. If not provided, the gene model track in the Variant Browser will be unavailable. 
    \item Any metadata associated with different accessions in the dataset, such as wild variety vs domesticated variety. \hl{This should be provided in a text file.}
    \item A GFF file of gene model information for the reference genome. 
    \item Whether there are prior identified introgressions known to be functionally relevant for the species. These can be used to `check' the ability of TersectBrowser+ to identify introgressed regions for the specified dataset.
\end{itemize}
    \item Clone the GitHub repository of TersectBrowser+ to the local machine. Follow the README from \hyperlink {https://github.com/Tersect-Browser/Tersect-browser.git}{TersectBrowser+ GitHub} for installation instructions.
    \item Download relevant files from section (1) to a new folder within the root directory of the local Tersect-browser repository.
    \item Set environment variables \verb+fasta+, \verb+gff+, and \verb+vcfs+ according to their new location relative to the root of Tersect-browser.
    \item Run the script \verb+setup_new_tbrowser_dataset.py+ using the environment variables of files as input:

@@

\subsubsection{User Interface}
\begin{figure}[h]
    \centering
    \includegraphics[width=10cm, height=8cm]{GenomeBrowser.png}
    \caption{UI diagram for Genome Browser.}
\end{figure}

\paragraph{Feature Description}
\begin{itemize}
    \item JBrowse window:
    \begin{itemize}
        \item Main window is located at the top of Tersect Browser, below settings bar but above the heatmap 
        \item Main window shows only sequence tracks and linear genome scale. Interactive features such as zoom anc chromosome selection have been removed from the main window. The container window has a dynamic offset, which changes according to tree view and zoom level to ensure the chromosome scales between TersectBrowser and JBrowse line up. 
        \item By default, the main window shows only the gene models track.
        \item Popup window provides a more sandbox experience, providing more user interaction. The user can change zoom level and displayed chromosome, and two scale bars enable both precise and rapid navigation. The hamburger menu on the top left provides the capability for the user to upload their own data to visualise unique tracks.
        \item By default, the popup window shows the reference sequence track, gene models track, and variant track for the specified accession.
    \item Tracks:
    \begin{itemize}
        \item Reference sequence 
        \item Variant tracks for each loaded accession
        \item Gene models FeatureTrack (\nameref{sec:Models})
    \end{itemize}
    \item Syncing of the JBrowse window and the TersectBrowser heatmap includes the following:
    \begin{itemize}
        \item Zoom - both by clicking the plus/minus buttons and by scrolling with the mouse wheel.
        \item Bin size.
        \item Selected interval.
        \item Chromosome.
        \item Horizontal scrolling along the x-axis.
    \end{itemize}
\end{itemize}
\end{itemize}

@@

The method \verb +JbrowserWrapper+ takes as arguments the props from \verb +JbrowseWrapperProps+ and defines the \verb +viewState+. If an accession has been selected and the accession name is stored under \verb +props.location.accession.name+, the \verb +viewState+ defined in the \verb +JbrowseWithAccessionName+ is returned (See below parameters). If not, a conditional check verifies whether the props \verb +defaultInterval+ and \verb +offsetCanvas+ have been populated. If either of them are empty, a container with the phrase ``Loading…" is returned. If all data is present, the \verb +viewState+ is defined using the function \verb +createViewState+. This takes as arguments the following variables: 
\begin{itemize}
\item \verb +assembly+: the assembly track 
\item \verb +tracks+: the data tracks to display 
\item \verb +configuration+: the config variable where the theme is defined 
\item \verb +defaultSession+: an object of type LinearGenomeView, which defines the initial state. This object includes the following configurations: 
\begin{itemize}
\item \verb +bpPerPx+: the base pairs per pixel, which defines the bin size 
\item \verb +assemblyName+: the corresponding assembly for the track 
\item \verb +start+: the track start position 
\item \verb +end+: the track end position 
\item \verb +refName+: the genomic coordinates of the viewing window 
\end{itemize}
\end{itemize}

@@

The configurations for the assembly track, which is built upon the reference accession SL2.50, is defined in \verb +assembly.ts+ in json format. The name is set to the reference accession name and \verb +trackId+ is set to ``SL2.50-ReferenceSequenceTrack". The paths to the fasta file, along with its corresponding fasta index, are specified as local server URLs provided by the backend during the Tersect Browser setup.

\paragraph{Tracks Config} 
The configurations for the \verb +VariantTrack+ and Gene Models \verb +FeatureTrack+ are defined in \verb +tracks.ts+ in json format. For the variant tracks, the \verb +name+ and \verb +trackId+ are set as the accession name, and for the Gene Models track these are set to ``ITAG2.4 Gene Models".  The paths to the zipped files, along with its corresponding Tabix index files, are specified as local server URLs provided by the backend during the Tersect Browser setup. Each track is separated by a comma.

\paragraph{Styling}
The container styling is defined in \verb +jbrowseConfig.ts+. The palette theme is set to colour `\#459e00' and the \verb +boxShadow+ is set to `none'.

@@

\subsubsection{Future Improvements}
The gene models track could be expanded to display alternative splicing events, offering a clearer view of transcript diversity. Additionally, adding a hover-based popup over each bin that displays gene models for accessions containing variants could offer a rapid and informative view of variant differences.

@@

\subsubsection{User Interface}
\begin{figure}[h]
    \centering
    \includegraphics[width=15cm, height=9cm]{FS1.png}
    \caption{UI digram for Feature Search.}
\end{figure}

\paragraph{Feature Description}
\begin{itemize}
    \item \verb+Open Variant Search+ button in TersectBrowser+ header, opens popup window for user to select advanced features.
    \item Popup window where user can select interval, choose impact level, and search button to initiate search.
    \item Output: Bins in the canvas are highlighted red at the chromosomal position where the variant is located, and only for accessions containing variants of the specified impact.
    \item Clear button located in TersectBrowser+ header to clear highlighted bins from canvas and restore original heatmmap.
\end{itemize}
\subsubsection{Technical Description}

\begin{figure}[h]
    \centering
    \includegraphics[width=7cm, height=15cm]{FS2.png}
    \caption{Data flow diagram for Feature Search.}
\end{figure}

\paragraph{\hl{Search Bar}}
\paragraph{\hl{Popup window with advanced settings }}
\paragraph{Highlighting bins}
Bin highlighting is controlled in the \verb+bin-draw.service+ by two functions. The first function, \verb+highlightFeatureBins()+, is defined, taking as arguments a string containing accession names, the bin position along the x-axis, and the \verb+binView+. First, the y-axis bin position for accession names is calculated. Then, the \verb+binView+ is redrawn in greyscale, with the bin colour determined by the difference to the reference accession. Using the x-axis \verb+binIndex+ position and the y-axis accession bin positions, these bins are coloured red in the \verb+binView+. The modified \verb+binView+ is returned as the output. The second function, \verb+highlightBins()+, takes as arguments the start position of the interval and the bin size currently shown in the Tersect pane, a list of ordered accessions shown in the canvas, and the searched accessions. For each searched accession, accession name is reformatted to match the format displayed in the Tersect pane. Then, the bin position along the x-axis is calculated using the \verb+getBinIndexFromPosition()+ function, which takes as arguments the feature position along the chromosome, the interval start position, and the bin size. Then, \verb+highlightFeatureBins()+ is called for each accession. Lastly, the canvas is redrawn using the modified \verb+binView+ output from \verb+highlightFeatureBins()+. 
\begin{itemize}
    \item highlightFeatureBins(accessions: string[], binIndex: number, binView: DistanceBinView) - takes a string of accession names, a binIndex (corresponding to the bin position along the x-axis matching the gene position on the chromosome), and binView. The y-axis index for bins matching accession names in the accessions strings is calculated and combined with the binIndex to colour these specific bins red. The rest of the bins are coloured in greyscale, with saturation depending on binDistance (calculated from tersect on the backend 
    \begin{itemize}
        \item Bin-draw.service.ts 
        \item Called by highlightBins() in bin-draw.service.ts 
    \end{itemize}
    \item highlightBins(intervalStart, binsize, orderedAccessions, searchedAccessions) - takes selected interval start position, selected binSize, list of ordered accessions shown in the canvas, and searched accessions (passed from callback function?). Accession names in Jbrowse are in a different format to what is stored in tersect browser, so accession names are reformatted to match tersect browser bins. binIndex is calculated for the selected gene. These two are passed to highlightFeatureBins() - along with this.bins - to highlight the bins. The canvas is then redrawn using the binView calculated in highlightFeatureBins(). 
    \begin{itemize}
        \item Bin-draw.service.ts 
        \item Called by callHighlightBins() in tersect-browser.component.ts 
    \end{itemize}
    \item callHighlightBins(searchedAcessions) - takes searchedAccessions. Calls highlightBins(), passing along  selected interval start position, selected binSize, list of ordered accessions shown in the canvas, and searched accessions (passed from callback function?). 
    \begin{itemize}
        \item Tersect-browser.component.ts 
    \end{itemize}
\end{itemize}

\paragraph{Calling Highlighting Bins} \mbox{}
\\
\hl{TBC: Mechanism} of searching VCFs to identify variants

Finally, callHighlightBins() is called, which itself calls highlightBins() from the bin-draw.service, passing as arguments the current interval start position, binsize, displayed accessions in an ordered format, and the searched accessions.

\paragraph{\hl{Clear Button }}
The clear button is located to the right of the \verb+Open variant search+ button and styled with \verb+fa fa-refresh+. It contains the event handler \verb+(click)="refreshBin()"+ that calls the \verb+refreshBin()+ method in \verb+BinDrawService+.
This method calls \verb+generatePlotArray()+, which re-creates the bin view, and \verb+updateCanvas()+, which redraws the canvas using the re-created bin view.

\subsubsection{Test Results}

\subsubsection{Future Improvements}
Currently, only bins on the TersectBrowser+ heatmap are highlighted in red to indicate the positions of high-impact variants. Highlighting could be extended across the Genome Browser and Gene Models extension, so that gene models matching the parameters of the feature search query are also highlighted in red.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=6cm, height=8cm]{Bar1.png}
    \caption{UI diagram for Barcode Generation.}
\end{figure}

\paragraph{Frontend Display Feature Description}
When the user clicks on a bin, the popup menu appears and the button ``Create barcode" is visible, styled with a barcode icon. When the user clicks on this button, a popup window appears in the centre of the screen. The window title is the corresponding accession name. The window contains the header ``Barcode Generator" and contains a description of what is being generated. There is a box for the user to input barcode size, with default set to 150, and the maximum number of variants allowed per barcode (this is an optional parameter. Leaving this blank means all possible barcodes will be returned).

At the bottom is a green ``Generate" button. Hovering over the button turns it dark green, and clicking on this button will call the backend scripts to generate the barcodes output. Whilst scripts are running and barcodes are being generated, a spinner is added to the button to indicate loading. After the script has been generated, the spinner disappears and another green button ``Download" appears at the bottom left of the window. Clicking on the button downloads the tsv file containing the barcodes.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=7cm, height=15cm]{Bar2.png}
    \caption{Data flow diagram for Barcode Generation.}
\end{figure}

@@

The method \verb+generateBarcode()+ calls and subscribes to the \verb+tersectBackendService generateBarcodes()+ method, setting the \verb+isLoading+ constant to ``True" and passing as arguments \verb+modalTitle+, \verb+chromosome+, \verb+startPosition+, \verb+endPosition+, \verb+barcodeSize+, and \verb+maxVariants+. The returned URL and filename are saved to the variables \verb+downloadUrl+ and \verb+downloadFileName+. Once returned, \verb+isLoading+ is set to ``False" and \verb+barcodeSize+ is set back to ``150".

The popup window is reset when closed by \verb+resetModal()+ on closing, which sets the \verb+barcodeSize+ to ``150``, \verb+maxVariants+ and \verb+downloadUrl+ to null, and \verb+downloadFileName+ to an empty string, removing the ``Download" button from the popup.

\paragraph{Modal Service Styling}:\\
The modal template is defined in \verb+global-barcode.component.html+. The popup window is visible depending on the modal state isVisible, and is centrally position with \verb+modalTitle+ set as the header. Explanation about the barcode service is described in the \verb+modal-title+ and \verb+modal-subtitle+ divs, and the \verb+input-container+ div contains input boxes for barcode size and maximum number of variants, the button ``Generate" which calls \verb+generateBarcode()+ and simultaneously shows a \verb+progressSpinner+ whilst the barcodes are being generated. An anchor displays a clickable ``Download" p-button that points towards the generated barcode file. The filename is set to \verb+downloadFileName+, and ``Download" is only visible once a \verb+downloadUrl+ has been generated for the barcodes file.

@@

The barcodes and respective metrics are written to a tsv file. The file name is formatted to include system date and time, \verb+`TB_Barcode_Gen'+, and the specified accession.

@@

\section {References}

\end{document}

>>>>

\author{David Oluwasusi, Tanya Stead, Gregory Lupton, Gabrielle Baumberg \\ Supervised by: Dr. Tomasz Kurowski}

@@

\section{Overview}
This Technical Documentation describes the particular details of building upon the original TersectBrowser developed by Tomasz Kurowski into  TersectBrowser+. It outlines the integration of multiple extension features, providing the user with additional display and data analysis capabilities.

\section{Architecture}
The TersectBrowser+ software is an Angular project, with an original frontend and backend written in Angular 8, and a separate extension space written in Angular 19 using Node 22. This 'bolt-on' approach was chosen for the extensions in order to seamlessly integrate the new extensions with the original workspace within the project time constrainsts. Additionally, this approach allows the new extnesions to use up-to-date packages, resulting in enhanced functionality and a modern design. To run TersectBrowser+ on different platforms, the new extensions are hosted on NPM and can be linked as dependencies within the project.

Extensions added to TersectBrowser+ include the integration of a genome browser capable of displaying a variety of different tracks, notably variants and gene models, search functionality to identify and display regions containing high impact variants, and a tool to generate accession-specific barcodes. These extensions are hosted in a separate section of the project, but interact with both frontent and backend services to provide a seamless user experience.

\section{Deployment}
The GitHub project \url{https://github.com/Tersect-Browser/Tersect-browser} README provides detailed instructions for setup, dependencies, and deployment. Here, we expand upon these instructions and give a detailed overview of the structure of each extension.

@@

The Admin is defined as an individual with bioinformatics and software design knowledge, background comprehension for all sections of this Technical Documentation, and acting to facilitate the use of a deployed TersectBrowser+ by the plant breeder main user of the software. When a new dataset is requested for deployment on TersectBrowser+, the Admin will need to follow the below steps:
\begin{enumerate}
    \item Collect background context and required data from the user.
\begin{itemize}
    \item Reference genome in fasta format (may be compressed). Size and chromosome number will impact the speed of dataset upload and deployment - TersectBrowser+ has been tested with a genome size of ~1GB, and 20 chromosome pairs.
    \item Resequenced genome dataset in VCF format (may be compressed, as a multi-sample VCF or as a directory of individual files). TersectBrowser+ has been tested with a dataset size of ~500 VCF files representing individual accessions. 
    \item Any metadata associated with different accessions in the dataset, such as wild variety vs domesticated variety. \hl{This should be provided in a text file.}
    \item A GFF file of gene model information for the reference genome, produced using SNPeff software. If not using SNPeff, ensure that variant impact levels are categorised into 'High', 'Medium', and 'Low' impact. If not provided, the gene model track in the Variant Browser will be unavailable. 
    \item Whether there are prior identified introgressions known to be functionally relevant for the species. These can be used to 'check' the ability of TersectBrowser+ to identify introgressed regions for the specified dataset.
\end{itemize}
    \item Clone the GitHub repository of TersectBrowser+ to the local machine. Follow the README from \hyperlink {https://github.com/Tersect-Browser/Tersect-browser.git}{TersectBrowser+ GitHub} for installation instructions.
    \item Download relevant files from section (1) to a new folder within the root directory of the local Tersect-browser repository.
    \item Set environment variables \verb+fasta+, \verb+gff+, and \verb+vcfs+ according to their new location relative to the root of Tersect-browser.
    \item Run the script \verb+setup_new_tbrowser_dataset.py+ using the environment variables of files as input:

@@

\paragraph{Feature Description}
\begin{itemize}
    \item JBrowse window:
    \begin{itemize}
        \item Main window is located at the top of Tersect Browser, below settings bar but above the heatmap 
        \item Main window shows only sequence tracks and linear genome scale. Interactive features such as zoom anc chromosome selection have been removed from the main window. The container window has a dynamic offset, which changes according to tree view and zoom level to ensure the chromosome scales between TersectBrowser and JBrowse line up. 
        \item By default, the main window shows only the gene models track.
        \item Popup window provides a more sandbox experience, providing more user interaction. The user can change zoom level and displayed chromosome, and two scale bars enable both precise and rapid navigation. The hamburger menu on the top left provides the capability for the user to upload their own data to visualise unique tracks.
        \item By default, the popup window shows the reference sequence track, gene models track, and variant track for the specified accession.
    \item Tracks:
    \begin{itemize}
        \item Reference sequence 
        \item Variant tracks for each loaded accession
        \item Gene models FeatureTrack (\nameref{sec:Models})
    \end{itemize}
    \item Syncing of the JBrowse window and the TersectBrowser heatmap includes the following:
    \begin{itemize}
        \item Zoom - both by clicking the plus/minus buttons and by scrolling with the mouse wheel.
        \item Bin size.
        \item Selected interval.
        \item Chromosome.
        \item Horizontal scrolling along the x-axis.
    \end{itemize}
\end{itemize}

@@

The method \verb +JbrowserWrapper+ takes as arguments the props from \verb +JbrowseWrapperProps+ and defines the \verb +viewState+. If an accession has been selected and the accession name is stored under \verb +props.location.accession.name+, the \verb +viewState+ defined in the \verb +JbrowseWithAccessionName+ is returned (See below parameters). If not, a conditional check verifies whether the props \verb +defaultInterval+ and \verb +offsetCanvas+ have been populated. If either of them are empty, a container with the phrase "Loading…" is returned. If all data is present, the \verb +viewState+ is defined using the function \verb +createViewState+. This takes as arguments the following variables: 
\begin{itemize}
\item \verb +assembly+: the assembly track 
\item \verb +tracks+: the data tracks to display 
\item \verb +configuration+: the config variable where the theme is defined 
\item \verb +defaultSession+: an object of type LinearGenomeView, which defines the initial state. This object includes the following configurations: 
\begin{itemize}
\item \verb +bpPerPx+: the base pairs per pixel, which defines the bin size 
\item \verb +assemblyName+: the corresponding assembly for the track 
\item \verb +start+: the track start position 
\item \verb +end+: the track end position 
\item \verb +refName+: the genomic coordinates of the viewing window 
\end{itemize}
\end{itemize}

@@

The configurations for the assembly track, which is built upon the reference accession SL2.50, is defined in \verb +assembly.ts+ in json format. The name is set to the reference accession name and \verb +trackId+ is set to "SL2.50-ReferenceSequenceTrack". The paths to the fasta file, along with its corresponding fasta index, are specified as local server URLs provided by the backend during the Tersect Browser setup.

\paragraph{Tracks Config} 
The configurations for the \verb +VariantTrack+ and Gene Models \verb +FeatureTrack+ are defined in \verb +tracks.ts+ in json format. For the variant tracks, the \verb +name+ and \verb +trackId+ are set as the accession name, and for the Gene Models track these are set to "ITAG2.4 Gene Models".  The paths to the zipped files, along with its corresponding Tabix index files, are specified as local server URLs provided by the backend during the Tersect Browser setup. Each track is separated by a comma.

\paragraph{Styling}
The container styling is defined in \verb +jbrowseConfig.ts+. The palette theme is set to colour '\#459e00' and the \verb +boxShadow+ is set to 'none'.

@@

\subsubsection{Future Improvements}

@@

\subsubsection{User Interface}
\paragraph{Feature Description}
\begin{itemize}
    \item \verb+Open Variant Search+ button in TersectBrowser+ header, opens popup window for user to select advanced features.
    \item Popup window where user can select interval, choose impact level, and search button to initiate search.
    \item Output: Bins in the canvas are highlighted red at the chromosomal position where the variant is located, and only for accessions containing variants of the specified impact.
    \item Clear button located in TersectBrowser+ header to clear highlighted bins from canvas and restore original heatmmap.
\end{itemize}
\subsection*{4.3.3 Feature Search Extension – Technical Description}

\subsubsection*{4.3.3.1 Creating Search Indexes Per Chromosome}

To enable searching for high impact variants across the dataset, distinct variant sites per chromosome region were obtained using \texttt{tersect}. The resulting VCF file was annotated with the merged VCF files from all accession samples in the dataset.

The final VCF was then compressed using \texttt{bgzip} and indexed with \texttt{tabix}. Additionally, Trix index files \citep{ucsc_trix} were generated for high impact variants on each chromosome to support rapid gene name search suggestions.

\subsubsection*{4.3.3.2 Handling Search Request}

An API endpoint was added to the \texttt{tersect-router} backend to support search functionality. The structure of the endpoint is as follows:

\begin{verbatim}
/query/:datasetId/variants/:chrom?start=number&end=number&filter=HIGH|MODERATE|LOW
\end{verbatim}

When searching by gene name, the Trix index files containing gene names associated with high impact variants are utilized.

\subsubsection*{4.3.3.3 Handling Searching by Interval}

The chromosome name is passed as a parameter to the search endpoint, while the start and end positions and the impact filter are passed as query parameters.

The search functionality is powered by \texttt{bcftools}, which is executed using the \texttt{child\_process} module in \texttt{Node.js}. The matching variants are captured by parsing their identifiers in the format \texttt{CHROM:POS:REF:ALT}, and these are stored in a set to ensure uniqueness.

To associate variants with accession names, a function uses \texttt{tersect} to determine which accessions have each specific variant. A function to get the optimal concurrency (\texttt(getOptimalConcurrency)) was introduced which detected the number of cores present on the server, subtracting 1 from the total and using the rest. To prevent server overload from concurrent \texttt{tersect} calls, the \texttt{p-limit} library is used to control concurrency based on available CPU cores.

The positions are then collated to a map that links positions to a list of accession names. The results are then constructed by looping through each line and getting specific accession list from the map created via the position.

The final result maps positions to accession name lists. These results are formatted and returned to the frontend in the following structure:

\begin{verbatim}
{
  region,
  filter,
  count,
  variants,
  totalAccessions
}
\end{verbatim}

The values are described below;

\begin{itemize}
    \item \textbf{region} - a string containing the region or interval on the selected chromosome (e.g., "chr1:100000-200000")
    \item \textbf{filter} - a string specifying the variant impact filter applied (e.g., "HIGH", "MODERATE", or "LOW")
    \item \textbf{count} - an integer representing the total number of variants found in the region with the specified filter
    \item \textbf{variants} - an array of variant objects, each including chromosome, position, reference, alternate, effect, and accessions
    \item \textbf{totalAccessions} - an integer specifying the total number of accessions in the dataset
\end{itemize}

Each variant is formatted as:

\begin{verbatim}
[
  {
    chr,
    pos: { start: pos },
    ref,
    alt,
    eff,
    accessions
  }
]
\end{verbatim}

The values are described below;

\begin{itemize}
    \item \textbf{chr} - a string specifying the chromosome name
    \item \textbf{pos} - an object with a \texttt{start} field indicating the variant position
    \item \textbf{ref} - a string specifying the reference allele
    \item \textbf{alt} - a string specifying the alternate allele
    \item \textbf{eff} - a string describing the effect or impact of the variant (e.g., "HIGH", "MODERATE", "LOW")
    \item \textbf{accessions} - an array of accession names (strings) that contain this variant
\end{itemize}

\subsubsection*{4.3.3.4 Handling Searching by Gene Name}

Gene-based searches use the pre-built Trix index files, which currently include only high impact variants.

The Trix index is generated by exporting high impact variants per chromosome to a text file, using this identifier format:

\begin{verbatim}
docId = "${chrom}:${pos}:${ref}:${alt}|${sample}|${effBlock}"
\end{verbatim}

The index is built using \texttt{ixIxx}.

\subsubsection*{4.3.3.5 Linking with the Frontend}

The gene search extension is developed in React to enable integration with future \texttt{JBrowse} sessions. Despite the main application being Angular (v19), React components are embedded using the \texttt{ReactComponentDirective} from the \texttt{@ngeenx/ngx-react} package.

Two UI components support search configuration:
\begin{itemize}
  \item A range slider to select intervals.
  \item An autocomplete input for gene ID search.
\end{itemize}

The React component handles the form submission, triggers the appropriate search via API or Trix, and returns results through a callback. Bin highlighting is then managed by the legacy tersect browser app.

\subsubsection*{4.3.3.6 Other Ways to Trigger Variant Search}

Variant searches can also be triggered from:
\begin{itemize}
  \item Each bin on the bin plot.
  \item A selected range on the scale plot.
\end{itemize}

These triggers emit a custom event \texttt{tersect-search-variants}, which sets the interval and opens the modal for user interaction.

\subsubsection*{4.3.3.7 Highlighting Bins}

Bin highlighting is handled by two main functions in \texttt{bin-draw.service.ts}:

\begin{itemize}
  \item \texttt{highlightFeatureBins(accessions, binIndex, binView)}: Computes the y-axis position for each accession, redraws the bin view in greyscale, and highlights selected bins in red.
  \item \texttt{highlightBins(intervalStart, binSize, orderedAccessions, searchedAccessions)}: Calculates the \texttt{binIndex} from position and bin size, and calls \texttt{highlightFeatureBins()}.
\end{itemize}

The top-level function \texttt{callHighlightBins(searchedAccessions)} is defined in \texttt{tersect-browser.component.ts}, where it retrieves interval, bin size, and accession info, and calls \texttt{highlightBins()}.

\subsubsection*{VCF Mechanism Summary}

The overall mechanism for variant search begins with API queries, uses either \texttt{bcftools} or Trix for data retrieval, links variants to accessions with \texttt{tersect}, and finally triggers the highlighting of bins in the Tersect UI.

\paragraph{Other Triggers for Variant Search}
Variant search also triggered via bin plot or scale plot selection using the \texttt{tersect-search-variants} custom event.

\clearpage

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=5cm, height=9cm]{UserBarcode.jpg}
    \caption{User-focused data flow through Barcoding Extension.}
\end{figure}
\paragraph{Frontend Display Feature Description}
When the user clicks on a bin, the popup menu appears and the button "Create barcode" is visible, styled with a barcode icon. When the user clicks on this button, a popup window appears in the centre of the screen. The window title is the corresponding accession name. The window contains the header "Barcode Generator" and contains a description of what is being generated. There is a box for the user to input barcode size, with default set to 150, and the maximum number of variants allowed per barcode (this is an optional parameter. Leaving this blank means all possible barcodes will be returned).

At the bottom is a green "Generate" button. Hovering over the button turns it dark green, and clicking on this button will call the backend scripts to generate the barcodes output. Whilst scripts are running and barcodes are being generated, a spinner is added to the button to indicate loading. After the script has been generated, the spinner disappears and another green button "Download" appears at the bottom left of the window. Clicking on the button downloads the tsv file containing the barcodes.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=7cm, height=12cm]{ExtBarcode.jpg}
    \caption{User-focused data flow through Barcoding Extension.}
\end{figure}

@@

The method \verb+generateBarcode()+ calls and subscribes to the \verb+tersectBackendService generateBarcodes()+ method, setting the \verb+isLoading+ constant to "True" and passing as arguments \verb+modalTitle+, \verb+chromosome+, \verb+startPosition+, \verb+endPosition+, \verb+barcodeSize+, and \verb+maxVariants+. The returned URL and filename are saved to the variables \verb+downloadUrl+ and \verb+downloadFileName+. Once returned, \verb+isLoading+ is set to "False" and \verb+barcodeSize+ is set back to "150".

The popup window is reset when closed by \verb+resetModal()+ on closing, which sets the \verb+barcodeSize+ to "150", \verb+maxVariants+ and \verb+downloadUrl+ to null, and \verb+downloadFileName+ to an empty string, removing the "Download" button from the popup.

\paragraph{Modal Service Styling}:\\
The modal template is defined in \verb+global-barcode.component.html+. The popup window is visible depending on the modal state isVisible, and is centrally position with \verb+modalTitle+ set as the header. Explanation about the barcode service is described in the \verb+modal-title+ and \verb+modal-subtitle+ divs, and the \verb+input-container+ div contains input boxes for barcode size and maximum number of variants, the button "Generate" which calls \verb+generateBarcode()+ and simultaneously shows a \verb+progressSpinner+ whilst the barcodes are being generated. An anchor displays a clickable "Download" p-button that points towards the generated barcode file. The filename is set to \verb+downloadFileName+, and "Download" is only visible once a \verb+downloadUrl+ has been generated for the barcodes file.

@@

The barcodes and respective metrics are written to a tsv file. The file name is formatted to include system date and time, \verb+'TB_Barcode_Gen'+, and the specified accession.

@@

\section {References}
\begin{thebibliography}{9}
\bibitem{ucsc_trix}
University of California, Santa Cruz. (n.d.). Trix Index Files. UCSC Genome Browser.
\end{thebibliography}
\end{document}
